---
title: "A View of Covid Between States"
author: "Herb"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 7
    fig_height: 12
    fig_caption: true
  word_document: default
  pdf_document: default
---

In the following assignment for our DataAnalysis class – I was trying to provide two visuals and an analysis of the Covid data we were provided. I did a few more as some just did not give much information.  I found setting up the data to be very time consuming, but I learned a lot.  My first two graphs were of California and just showed how covid increased overtime by administration site acummulative reporting data--i.e. nothing really surprising -- so I moved on to look at the data at the start of 2020 for each state — so I could do a state by state comparison.  Rank ordered the states by death count at the end of the data reporting period by aggregating all the reports per state.

Relooking at the data, I realized I needed to adjust for the population per state to provide information on which state had more or less deaths per person so to speak (this is done in most accident reports or medical reposts for the same reason). So I tallied the states by admin reports and used one of the last cummulative date counts provided.  I then compared population totals with those found online and everything look copacetic.

To show the dramatic change in which state has the highest death rate versus just high numbers of deaths per state, I provided the first graph of JUST total death counts--then keep the ordering for the death ratio plots.  The values were all over the place, showing that the states had different death rates than population rates.  I followed this with regression of both variables and deaths per population were not significant but death by state temperature was.  Take a look below.

Finally out of curiosity I through in some comparisons with smokers, pets, and temperature per state--which was fun.  I would love to have delved further into the climate, financial, population density, altitude average and other variables across the states.  Too fun. When you check the death ratio per state to the population rate there is NO correlation.  However, the second you look at the ratio to other predictors such as state temperature, the correlation climbs significantly.  So there are variables or predictors out there that can model how the states did once the population effect is controlled for.  Research will be needed to answer further questions.

In relation to biases--there are many.  Just by my looking at smoking rates per state and finding a correlation is due somewhat to bias.  There are many other variables that can be attributed to smoking that makes it correlated with the death rate--as you see in my quick correlation analyzed, pet ownwership per state was the highly correlated with reduce deaths while smoking was correlated with increased deaths (as you would expect with a lung desease).

```{r library in}

library(tidyverse)
library(dplyr)
library(readxl)
library(writexl)
library(devtools)
library(ggplot2)
library(markdown)
library(lubridate)
library(RCurl)
library(knitr)
suppressWarnings({ 
 # Code that generates warning messages 
}) 
```
```{r readin}

url_in<-"https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

```

In the follwoing code chucks I combine the two US Death databases we downloaded.  Removed all. the time data except the 2020 end point count per precinct and state.  I then combined the precincts per each state to get the total reported deaths at the end of January 2020.

Note: I left my first graph of the points of deaths in the database which alert me to the fact that multiple administration points were reporting the tallys of deaths.  this enabled me to add the data together to get a more meaningful chart and exminationi of death BY state.

Obviously, a more detailed analysis can be achieved by looking at admin counts and location and population per thos locationis but that was not my intent in this investigatioini -- besides it took me long enough too learn how to do a smaller data base, clean its errors and graphs its visualiztions -- maybe next time I will dig deeper.

```{r}

file_names<-c("time_series_covid19_confirmed_US.csv",  "time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_US.csv",  "time_series_covid19_deaths_global.csv")
urls<-str_c(url_in,file_names)
us_cases<-read_csv(urls[1])
us_deaths<-read_csv(urls[3])
head(us_cases)
```

```{r}
us_deaths1 = us_deaths[,-c(1:5,8,11)]

deaths <- us_deaths1 %>%
  pivot_longer(
    cols = -c(Admin2, Province_State, Lat, Long_, Population),
    names_to = "date" ,
    values_to = "deaths") %>%
    mutate(date=mdy(date))
```

```{r}
head(deaths)
```

```{r}
head(us_cases)
```

```{r}
us_cases1 = us_cases[,-c(1:5,8,11)]

cases <- us_cases1 %>%
  pivot_longer(cols = -c(Admin2, Province_State, Lat, Long_),
    names_to = "date" ,
    values_to = "cases") %>%
    mutate(date=mdy(date))
head(cases)
```

```{r}
cases[,6]
usdata = cbind(deaths,cases[,6])
head(usdata)
tail(usdata)
sum(is.na(usdata$cases))
sum(is.na(usdata$deaths))
```
```{r}
cal = usdata[usdata$Province_State == "California",]
head(cal)
```
```{r}
p <- ggplot(cal, aes(x=date, y=deaths)) +
  geom_point() + 
  labs(x="Date",
  y = "Deaths",
  title = "California Deaths by Dates Given")
p
```
#This was my first plot which was done out of curiousity of what was in there.  I did learn a lot.  Data wwas reported by admin within the state which had to be added toegther to get an overal state average.  The curves confirmed that the reports were cumulative overtime and I was not losing any end count data by looking at the last non NA data reported per admin area.


```{r}
yearlycases = us_cases[,c(10,1104)]
head(yearlycases)
yearlydeaths = us_deaths[,c(6,7, 9,10,11,12,1105)]
head(yearlydeaths)
```

```{r}
usdata = cbind(yearlydeaths,yearlycases)
head(usdata)
tail(usdata)
sum(is.na(usdata$cases))
sum(is.na(usdata$deaths))
```

```{r}
#checking which values are not NA 
summary(usdata)
```
```
The following chunk is the key to simplifying the data so I could compare state to state efficiency so to speak.  I utillized the "aggregate" function and then check numbers to what was report by CDC -- they were darn close


```{r}
sum(usdata$deaths, na.rm = TRUE)
usdata[usdata$"Population">1000000,]
colnames(usdata) = c("city", "state", "lat", "long", "city/state","population", "deaths", "longtocheck", "cases")
head(usdata)
```

```{r}
#. aggregate the data bystate summing deaths and cases and taking mean of population.
statecases=aggregate(usdata$cases, list(usdata$state), FUN=sum)
statedeaths=aggregate(usdata$deaths, list(usdata$state), FUN=sum)
statepop=aggregate(usdata$population, list(usdata$state), FUN=sum)
head(statecases)
head(statedeaths)
head(statepop)
```

```{r}

bystate =data.frame(statepop,statecases$x, statedeaths$x)
names(bystate)[1] = "state"
names(bystate)[2] = "population"
names(bystate)[3] = "cases"
names(bystate)[4] = "deaths"
bystate[1:10,]
```

```{r}

bystate = na.omit(bystate)
bystate
plot(bystate$cases,bystate$deaths)
summary(bystate)
bystate
bystatesort = bystate[order(bystate$death, decreasing = TRUE),]
print(bystatesort)
```

```{r fig.height = 6}


barplot(bystate$deaths ~ bystate$state, horiz = TRUE)
```

```{r fig.height = 6}

plotpop = bystatesort %>%
  ggplot() +
labs(title = "Pop by State",
       x = "Pop",
       y = "State") + 
  geom_bar(aes(x = reorder(state, population), y = population, 
               fill = state), stat = "identity", show.legend = FALSE) +
 
  coord_flip()
plotpop
```


```{r fig.height = 6}
plotcases = bystatesort %>%
  ggplot() +
  labs(title = "Cases by State",
       x = "Cases",
       y = "State") + 
  geom_bar(aes(x = reorder(state, population), y = cases, 
               fill = state), stat = "identity", show.legend = FALSE) +
  coord_flip()
plotcases
```

```{r fig.height = 6}
plotdeaths = bystatesort %>%
  ggplot() +
  labs(title = "Deaths by State",
       x = "Deaths",
       y = "State") + 
  geom_bar(aes(x = reorder(state, population), y = deaths, 
               fill = state), stat = "identity", show.legend = FALSE) +
  coord_flip()
plotdeaths
```

```{r fig.height = 6}

plotdeaths = bystatesort %>%
  ggplot() +
  labs(title = "Deaths per Polulation by State",
       x = "state",
       y = "Deaths/Population") + 
  geom_bar(aes(x = reorder(state, population), y = deaths/population, 
               fill = state), stat = "identity", show.legend = FALSE) +
  coord_flip()
plotdeaths
```
#I had to convert my collected data because I had trouble linking it in git..I will learn to do it later.
# temp data	from:	https://wisevoter.com/state-rankings/average-temperature-by-state/		
# hospital bed data		https://ceoworld.biz/2020/03/16/these-are-the-u-s-states-with-the-most-and-least-hospital-beds/		
# Centers for Disease Control and Prevention. Behavioral Risk Factor Surveillance System 2017, analysed by the American Lung Association Epidemiology and Statistics Unit
# SMOKING DATA FROM https://www.statista.com/statistics/261595/us-states-with-highest-smoking-rates-among-adults/	
```{r NEWDATA}
state = c("AL","AK","AZ","AR","AZ","CA","CP","DE","FL","GA","GU","ID","IL","IN","IS","KS","KY","LA","ME","AMD","MA","MI","MO","MT","NE","ANV","NH","NJ","NM","NY","NC","ND","OH","OK","OR","PA","RI","SC","SD","TN","TX","UT","VT","VA","VI","WA","WV","WI","WY")

pop_in_thousands = c(4903.185,	740.995,	7278.717,	3017.804,	39512.223,	5758.736,	3565.287,	973.764,	21477.737,	10617.423,	1787.065,	12671.821,	6732.219,	3155.07,	2913.314,	4467.673,	4648.794,	1344.212,	6045.68,	6892.503,	9986.857,	5639.632,	2976.149,	6626.371,	1068.778,	1934.408,	3080.156,	1359.711,	8882.19,	2096.829,	19453.561,	10488.084,	762.062,	11689.1,	3956.971,	4217.737,	12801.989,	1059.361,	5148.714,	884.659,	6829.174,	28995.881,	3205.958,	623.989,	8535.519,	7614.893,	1792.147,	5822.434,	578.759)

deaths_per_pop = c(4.25152222483957,	1.96357600253713,	4.48307029933984,	4.2302283382221,	2.51393094233144,	2.42848430627832,	3.33633729907298,	3.30675605177435,	3.91922109857291,	3.93428800943506,	2.99037807802178,	3.2339471966973,	3.82073132202027,	3.34002098210182,	3.39922164243195,	3.98261018655573,	3.97500943255391,	2.12243306859335,	2.67232139312699,	3.37453607201912,	4.12392006814556,	2.55708173866664,	4.41879758036308,	3.39401461222138,	3.39640224630372,	2.44519253435676,	3.84201319673419,	2.13868976569286,	4.01916644431159,	4.24545826102176,	3.90226755913737,	2.66654996279587,	3.18609247016647,	3.55288260002909,	4.42308017925833,	2.16727595864797,	3.8585410438956,	3.58518012273437,	3.68694007862934,	3.55504211227151,	4.22496190608117,	3.17693399279711,	1.62884229924409,	1.41669164039751,	2.69017033410622,	2.0107964747502,	4.3467416456351,	2.74610240322175,	3.40383475678132)

temp_rank = c(6,	49,	8,	9,	11,	37,	28,	15,	1,	4,	42,	22,	23,	33,	18,	14,	2,	44,	17,	30,	40,	47,	5,	19,	45,	29,	27,	39,	21,	20,	36,	12,	48,	25,	10,	34,	31,	26,	7,	38,	13,	3,	32,	41,	16,	35,	24,	43,	46)

medbed_per_thou = c(3.1,	2.2,	1.9,	3.2,	1.8,	1.9,	2,	2.2,	2.6,	2.4,	1.9,	2.5,	2.7,	3,	3.3,	3.2,	3.3,	2.5,	1.9,	2.3,	2.5,	2.5,	4,	3.1,	3.3,	3.6,	2.1,	2.1,	2.4,	1.8,	2.7,	2.1,	4.3,	2.8,	2.8,	1.6,	2.9,	2.1,	2.4,	4.8,	2.9,	2.3,	1.8,	2.1,	2.1,	1.7,	3.8,	2.1,	3.5)

smoker_rate = c(14,	14,	11,	17,	9,	10,	9,	12,	10,	12,	11,	12,	15,	4,	13,	17,	15,	12,	9,	9,	14,	11,	15,	15,	13,	13,	14,	9,	9,	14,	10,	13,	13,	15,	15,	12,	13,	10,	13,	13,	17,	11,	9,	11,	11,	9,	20,	13,	14)

pet_own_rate = c(59.8,	59.3,	58,	69,	57.2,	64.7,	49.9,	57.9,	56,	51.1,	69.9,	48.6,	69.2,	59.4,	62.8,	64.1,	54.4,	63.5,	48.6,	49.1,	62.4,	54,	65.5,	63.5,	61.9,	70.3,	53.3,	51.8,	47.4,	60.1,	49.7,	58.6,	63.7,	62.4,	65,	59.2,	60.6,	45.4,	62,	46.4,	61.7,	58.2,	58.5,	70,	55.5,	62.7,	70.7,	59,	71.8)


my_data <- data.frame(pop_in_thousands,deaths_per_pop, temp_rank, medbed_per_thou, smoker_rate, pet_own_rate)

summary(my_data)
```


```{r fig.height = 6}

# Function to add correlation coefficients
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...) {
    usr <- par("usr")
    on.exit(par(usr))
    par(usr = c(0, 1, 0, 1))
    Cor <- abs(cor(x, y)) # Remove abs function if desired
    txt <- paste0(prefix, format(c(Cor, 0.123456789), digits = digits)[1])
    if(missing(cex.cor)) {
        cex.cor <- 0.4 / strwidth(txt)
    }
    text(0.5, 0.5, txt,
         cex = 1 + cex.cor * Cor) # Resize the text by level of correlation
}
```

```{r}
# Plotting the correlation matrix

pairs(my_data,
      upper.panel = panel.cor,    # Correlation panel
      lower.panel = panel.smooth) # Smoothed regression lines
```

I found this one interesting.  

```{r multiple}

model = lm(deaths_per_pop ~ temp_rank + medbed_per_thou + smoker_rate + pet_own_rate)
summary(model)
```
##Just a quick check to see if there was anything to the random variables I choose.  THe main thing I took from this is that the state temperature is likely the highes predictor of deaths.  And sure enough if we jsut do the one variable it gives an R sqyuat of 67% not bad for a cold day in May.  So based on the model the colder the state the LESS Deaths -- hum the virus didn't like the cold -- or people stayed home - or ... the list could go on.  Smoking and pets did still look a bit promising.  Weird huh?
```{r for rank}

mydatalm = lm(deaths_per_pop ~ temp_rank, data = my_data)
summary(mydatalm)
```

```{r}
my_data1 <- data.frame(state, pop_in_thousands,deaths_per_pop, temp_rank, medbed_per_thou, smoker_rate, pet_own_rate)
summary(my_data1)
```


```{r}
ggplot(data = my_data1, aes(y = deaths_per_pop, x = temp_rank)) +
        geom_point(aes(color = "red")) +
        geom_smooth(method = "lm") +
        geom_text(aes(label=ifelse(temp_rank<7,as.character(state),'')),hjust=0,vjust=2) +
        geom_text(aes(label=ifelse(temp_rank>45,as.character(state),'')),hjust=0,vjust=0) +
        labs(title = "Scatterplot of deaths per unit of population versus state temp ranking",
             y = "Deaths/Population)",
             x = "Ranking - #1 warm -- 50 is coldest")
```

```{r for population}
mydatalm = lm(deaths_per_pop ~ pop_in_thousands, data = my_data)
summary(mydatalm)
```
```{r}
ggplot(data = my_data1, aes(y = deaths_per_pop, x = pop_in_thousands)) +
        geom_point(aes(color = "red")) +
        geom_smooth(method = "lm") +
        geom_text(aes(label=ifelse(temp_rank<7,as.character(state),'')),hjust=0,vjust=2) +
        geom_text(aes(label=ifelse(temp_rank>45,as.character(state),'')),hjust=0,vjust=0) +
        labs(title = "Scatterplot of deaths per unit of population versus state population/1000",
             y = "Deaths/Population)",
             x = "Stat pop/1000")
```
##In conclusion, or recap, I started the exploration as started in class then decided to look into DID OUR STATES VARY IN COVID DEATHS.  the first look was always California had the most daeaths.  Made me feel bad as this is home for now.  So after sizing the data down to the total count of deaths by the year 2020 and adjusting it to cover the ratio of deaths per population--taking the shear number issue out of the equation.  And sure enough the death rate ratios were all over the place.  So a gleamed so data for other sources such as smoking, temperature, pet ownership (my favorite -- as my pups saved me during our lock down).  And sure enough starte temp had the highest correlation to death rate ratio...opposite of what I would have expected.  The lowest death ratio  states were the coldest states out there.  We can have some fun trying to figure out why--altitude, pop density, snow drift removal exercise, etc....I will leave that to another class or maybe my next course and some more downlads from the CDC.

#All in all I enjoyed our journey to explore how to manipulate a database and pull sosme visualizations/models from it. I found the tools covered in the classes amazing.  I have a lot to learn but enjoy what we have covered so far.  Thank you to all the staff and your help.  Herb


